---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
# Read the csv file
activity <- read.csv("activity.csv",header=TRUE)
# Adjust the date
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?
```{r}
# Calculate total number of steps taken on each day
total_steps_per_day <- as.integer(with(activity,tapply(steps,date,sum,na.rm=TRUE)))
hist(total_steps_per_day,breaks=10,col="gray",
     xlab="Number of steps per day", main = "Histogram of total steps taken per day")
cat("Mean of total steps per day : ", mean(total_steps_per_day))
cat("Median of total steps per day : ",median(total_steps_per_day))
```

## What is the average daily activity pattern?
```{r}
# Average daily activity pattern
total_steps_per_interval <- with(activity,tapply(steps,interval,sum,na.rm=TRUE))
plot(names(total_steps_per_interval),total_steps_per_interval,type="l",
     xlab="Interval",ylab="Average steps",
     main="Average steps across all days per 5 minute interval")
cat("Maximum average steps across all days occur at inteval : ", names(total_steps_per_interval)[which.max(total_steps_per_interval)])
```


## Imputing missing values
```{r}
cat("Number of missing values : ", sum(is.na(activity$steps)))
```

Check how the NA values are distributed on a day basis. Note each day has at max 288 entries
```{r}
with(activity,tapply(steps,date,function(x) {sum(is.na(x))}))
```
Since NA values on these specific days correspond to data not being available in any of the intervals, any sort of imputing isn't going to change the results.

In other words, when data is missing on a specific day, it is missing across all the intervals of that day thereby any imputing based on mean/median would still yeild a NA result.

Check how NA values are distributed on an interval basis. Note each interval has at max 61 days of data
```{r}
with(activity,tapply(steps,interval,function(x) {sum(is.na(x))}))
```
Bingo. This is the way to go. Pick mean of remaining data across other days in that interval and asssing it to missing intervals
```{r}
# Substitue missing values with mean value of that interval
new_activity <- activity
mean_steps_per_interval <- with(activity,tapply(steps,interval,mean,na.rm=TRUE))
U <- which(is.na(new_activity$steps))
new_activity$steps[U] <- mean_steps_per_interval[as.character(new_activity$interval[U])]
```
Done imputing the data. Missing values are replaced with mean steps of the interval in which the current NA value belongs.
```{r}
#histogram
total_steps_per_day_new <- as.integer(with(new_activity,tapply(steps,date,sum,na.rm=TRUE)))
hist(total_steps_per_day_new,breaks=10,col="gray",
     xlab="Number of steps per day", 
     main = "Histogram of total steps taken per day - after Imputing")
cat("Mean of total steps per day (after imputing): ", mean(total_steps_per_day_new))
cat("Median of total steps per day (after imputing): ",median(total_steps_per_day_new))
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
U <- weekdays(new_activity$date) %in% c("Saturday","Sunday")
new_activity$day <- NA
new_activity$day[U] <- as.character("Weekend")
new_activity$day[!U] <- as.character("Weekday")

weekday <- new_activity[new_activity$day=="Weekday",]
weekday_mean_steps <- with(weekday,tapply(steps,interval,mean,na.rm=TRUE))

weekend <- new_activity[new_activity$day=="Weekend",]
weekend_mean_steps <- with(weekend,tapply(steps,interval,mean,na.rm=TRUE))

par(mfrow=c(2,1),mar=c(2,3,2,1))
plot(names(weekday_mean_steps),weekday_mean_steps,type="l",xlab="Interval", main="Average number of steps")
legend("topright",legend="Weekday")
plot(names(weekend_mean_steps),weekend_mean_steps,type="l",xlab="Interval")
legend("topright",legend="Weekend")
```